---
title: "Final Covid 19"
author: "A. Ayzenberg"
date: "2025-04-28"
output: 
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Necessary Libraries

```{r libraries}
library(tidyverse)
library(lubridate)
library(ggplot2)
```


## Import the data
```{r import}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")
urls <- str_c(url_in, file_names)
global_cases <- read_csv(urls[2])
global_deaths <- read_csv(urls[4])
US_cases <- US_cases <- read_csv(urls[1])
US_deaths <- read_csv(urls[3])
```

## Transform the data for visualization and analysis

```{r tidying}
# transform global data
global_cases <- global_cases %>% pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), names_to = "date", values_to = "cases") %>% select(-c(Lat, Long))
global_deaths <- global_deaths %>% pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), names_to = "date", values_to = "deaths") %>% select(-c(Lat, Long))
global <- global_cases %>% full_join(global_deaths) %>% rename(Country_Region = `Country/Region`, Province_State = `Province/State`) %>% mutate(date = mdy(date))
global <- global %>% filter(cases > 0)
summary(global)
# Verify that the following isn't a single outlier
global %>% filter(cases>100000000)
# transform US data
US_cases <- US_cases %>% pivot_longer(cols = -(UID:Combined_Key), names_to = "date", values_to = "cases") %>% select(Admin2:cases) %>% mutate(date = mdy(date)) %>% select(-c(Lat, Long_))
US_deaths <- US_deaths %>% pivot_longer(cols = -(UID:Population), names_to = "date", values_to = "deaths") %>% select(Admin2:deaths) %>% mutate(date = mdy(date)) %>% select(-c(Lat, Long_))
US <- US_cases %>% full_join(US_deaths)
global <- global %>% unite("Combined_Key", c(Province_State, Country_Region), sep = ", ", na.rm = TRUE, remove = FALSE)
# add population statistics to global data
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>% select(-c(Lat,Long_, Combined_Key, code3, iso2, iso3, Admin2))
global <- global %>% left_join(uid, by = c("Province_State", "Country_Region")) %>% select(-c(UID, FIPS)) %>% select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
# Final preparations for plotting and visualizing the data
US_by_state <- US %>% group_by(Province_State, Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths *1000000 / Population) %>% select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()
US_totals <- US_by_state %>% group_by(Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths *1000000 / Population) %>% select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()
```

## Visualizing US statistics

```{r visualize_US}
# Find out maximum date (how far the data goes)
max(US_totals$date)
US_totals %>% filter(cases > 0) %>% ggplot(aes(x = date, y = cases)) + geom_line(aes(color = "cases")) + geom_point(aes(color = "cases")) + geom_line(aes(y = deaths, color = "deaths")) + geom_point(aes(y = deaths, color = "deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + labs(title = "COVID19 in US", y= NULL)
```

## Visualizing data for Colorado

```{r visualize_CO}
state <- "Colorado"
US_by_state %>% filter(Province_State == state) %>% filter(cases > 0) %>% ggplot(aes(x = date, y = cases)) + geom_line(aes(color = "cases")) + geom_point(aes(color = "cases")) + geom_line(aes(y = deaths, color = "deaths")) + geom_point(aes(y = deaths, color = "deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + labs(title = str_c("COVID19 in ", state), y= NULL)
```

## Evaluate only new cases
In this case only new cases will be looked at to visualize if there is a trend for US data

```{r analysis_US}
## Transforming and looking at US new cases
US_by_state <- US_by_state %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))
US_totals %>% ggplot(aes(x = date, y = new_cases)) + geom_line(aes(color = "new_cases")) + geom_point(aes(color = "new_cases")) + geom_line(aes(y = new_deaths, color = "new_deaths")) + geom_point(aes(y = new_deaths, color = "new_deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + labs(title = "COVID19 in US", y= NULL)
US_state_totals <- US_by_state %>% group_by(Province_State) %>% summarize(deaths = max(deaths), cases = max(cases), population = max(Population), cases_per_thou = 1000 * cases / population, deaths_per_thou = 1000 * deaths / population) %>% filter(cases > 0, population > 0)
US_state_totals %>% slice_min(deaths_per_thou, n = 10)
```

## Evaluate only new cases (state)
In this case only new cases will be looked at to visualize if there is a trend for CO data

```{r analysis_CO}
## Transforming and looking at Colorado new cases
US_by_state %>% filter(Province_State == state) %>% ggplot(aes(x = date, y = new_cases)) + geom_line(aes(color = "new_cases")) + geom_point(aes(color = "new_cases")) + geom_line(aes(y = new_deaths, color = "new_deaths")) + geom_point(aes(y = new_deaths, color = "new_deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + labs(title = str_c("COVID19 in ", state), y= NULL)
US_state_totals <- US_by_state %>% group_by(Province_State) %>% summarize(deaths = max(deaths), cases = max(cases), population = max(Population), cases_per_thou = 1000 * cases / population, deaths_per_thou = 1000 * deaths / population) %>% filter(cases > 0, population > 0)
US_state_totals %>% slice_min(deaths_per_thou, n = 10)
```

## Worst and Least impacted
What if the goal is to see which states suffered the least deaths, the most?

```{r faired_best}
## Who faired the best, the worst?
US_state_totals <- US_by_state %>% group_by(Province_State) %>% summarize(deaths = max(deaths), cases = max(cases), population = max(Population), cases_per_thou = 1000 * cases / population, deaths_per_thou = 1000 * deaths / population) %>% filter(cases > 0, population > 0)
US_state_totals %>% slice_min(deaths_per_thou, n = 10)
US_state_totals %>% slice_max(deaths_per_thou, n = 10)
```

## Modeling the data
Below the data will be plotted to determine a relation between cases per thousand and deaths per thousand, and also visualize to see which points faired better than expected and which did worse.

```{r modeling}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)
US_state_totals %>% slice_min(cases_per_thou)
US_state_totals %>% slice_max(cases_per_thou)
x_grid <- seq(150, 450)
US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))
US_tot_w_pred %>% ggplot() + geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") + geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```


## Potential Biases
The biggest source of bias in this case would be reporting or rather under reporting in some parts of the world, especially in the early days of covid, most notably due to lack of funding, insufficient numbers of tests and potential for remote areas. The same can be said for deaths as one factor not put into consideration is a country or region's wealth, meaning a more accurate comparison would be between 2 regions of comparable wealth (i.e. GDP per capita) as opposed to already striggling countries being labeled as handling the epidemic worse than regions with higher fund availability. 